---
title: 'ggplot2: Scales'
author: Fei
date: '2021-03-30'
slug: ggplot2-scales
categories:
  - R
tags:
  - plot
  - R
weight: 1
draft: no
output:
  blogdown::html_page:
    toc: yes
    toc_depth: 3
    number_sections: yes
---

```{r include=FALSE}
knitr::opts_chunk$set(warning = F, message = F, cache = TRUE,tidy=F,fig.align='center',fig.showtext=TRUE,results="hold",fig.show = "hold")
library(tidyverse)
library(patchwork)
```

# Introduction
The scales toolbox divide scales into three main groups:

* Position scales and axes.
* Color scales and legends.
* Scales for other aesthetics.

# Position scales and axes
Sometimes an aesthetic is mapped to a **computed variable** that does not need to be explicitly specified, as happens with `geom_histogram()`, which use the computed variable `count`.

```{r eval=FALSE, include=FALSE}
ggplot(mpg, aes(x = displ)) + 
  geom_histogram()
ggplot(mpg, aes(x = displ, 
                y = after_stat(count))) + 
  geom_histogram()
```

## Numeric
The most common continuous position scales are the default `scale_x_continuous()` and `scale_y_continuous()`. Other functions like `scale_x_log10()`, `scale_x_reverse()` transform the data, which can also be achieved by setting the `trans` option.

| Name      | Function                | Name      | Function     
|:----------|:------------------------|:----------|:-----------------------
| asn       | $\tanh^{-1}(x)$         | logit     | $\log(\frac{x}{1 - x})$ 
| exp       | $e ^ x$                 | pow10     | $10^x$                  
| identity  | $x$                     | probit    | $\Phi(x)$               
| log       | $\log(x)$               | reciprocal| $x^{-1}$                
| log10     | $\log_{10}(x)$          | reverse   | $-x$                    
| log2      | $\log_2(x)$             | sqrt      | $x^{1/2}$               

Specifically, if you use a transformed scale, the axes will be labeled in the original data space; if you transform the data, the axes will be labeled in the transformed space. To transform after statistical computation use `coord_trans()`.

```{r fig.height=2, fig.width=6}
# manual transformation
ggplot(mpg, aes(log10(displ), hwy)) + 
  geom_point() |

# transform using scales
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  scale_x_log10()
```

By default, ggplot2 converts data outside the scale limits to `NA`, because the default value of `oob` (out of boundary) argument is `scales::censor()`. Setting it to `scales::squish()` will squishes all values into the range.

The space around the plot can be eliminated with `expand = c(0, 0)`. To make sure some values are shown on the plots, use `expand_limits()`.

```{r fig.height=2, fig.width=5}
ggplot(faithfuld, aes(waiting, eruptions)) + 
  geom_raster(aes(fill = density)) + 
  theme(legend.position = "none") +
  expand_limits(x=1,y=1)|

ggplot(faithfuld, aes(waiting, eruptions)) + 
  geom_raster(aes(fill = density)) + 
  scale_x_continuous(expand = c(0,0),
                     trans = "reverse") + 
  scale_y_continuous(expand = c(0,0)) +
  theme(legend.position = "none")|
  
ggplot(faithfuld, aes(waiting, eruptions)) + 
  geom_raster(aes(fill = density)) +
  # multiply 5
  scale_x_continuous(expand = c(5,0)) + 
  # add 5 at each side
  scale_y_continuous(expand = c(0,5)) +
  theme(legend.position = "none")
```

Breaks can be added manually, but ggplot2 also allows you to pass a function to `breaks`:

* `scales::breaks_extended()`	creates automatic breaks for numeric axes (the standard method).
* `scales::breaks_log()`	creates breaks appropriate for log axes.
* `scales::breaks_pretty()`	creates "pretty" breaks for date/times.
* `scales::breaks_width()`	creates equally spaced 

Every break is associated with a label and these can be changed by setting the `labels` argument to the scale function:

* `scales::label_bytes()` formats numbers as kilobytes, megabytes etc.
* `scales::label_comma()` formats numbers as decimals with commas added.
* `scales::label_dollar()` formats numbers as currency.
* `scales::label_ordinal()` formats numbers in rank order: 1st, 2nd, 3rd etc.
* `scales::label_percent()` formats numbers as percentages.
* `scales::label_pvalue()` formats numbers as p-values: <.05, <.01, .34, etc.

As with breaks, you can also supply a function to minor_breaks, such as `scales::minor_breaks_n()` or `scales::minor_breaks_width()` functions that can be helpful in controlling the minor breaks. However, these minor breaks do not add ticks on the axis.

```{r fig.height=2, fig.width=9}
toy <- data.frame(
  const = 1,   up = 1:4,  txt = letters[1:4], 
  big = (1:4)*1000,  log = c(2, 5, 10, 2000))
# add breaks without labels
brks = seq(500, 4000, 250)
lbs <- ifelse(brks%%1000 == 0, brks, "")
# plot
axs <- ggplot(toy, aes(big, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL)
# minor breaks
axs + 
  scale_x_continuous(breaks = scales::breaks_extended(n = 4),
                         minor_breaks = seq(1500,4000,250))|
# use offset to control breaks
axs + 
  scale_x_continuous(breaks = scales::breaks_width(800, offset = 200))|
# control labels
axs + 
  scale_y_continuous(labels = scales::label_dollar(prefix = "￥", 
                                                   suffix = "€"))|
# breaks without labels
axs + 
  scale_x_continuous(breaks = brks, labels = lbs)
```

## Date-time
Different formats of date/time can be converted by a few functions (`as.Date()`, `as.POSIXct()` or `hms::as_hms()`) or the **lubridate** package. Default scale for dates is `scale_x_date()` and `scale_x_datetime()` is default for data-time data.

Breaks and minor breaks can be specified by `date_breaks` and `data_minor_breaks` respectively.  Date label formatting is useful when there's no sufficient room.

```{r fig.height=2, fig.width=6}
date_base <- ggplot(economics, aes(date, psavert)) + 
  geom_line(na.rm = TRUE) +
  labs(x = NULL, y = NULL)

date_base + scale_x_date(date_breaks = "25 years",
                         date_minor_breaks = "5 years",
                         date_labels = "%y") |
# formatting with function label_date, or label_date_short
date_base + scale_x_date(labels = scales::label_date("%Y"))
```

## Discrete
The default scales mapping discrete variables to position scales are `scale_x_discrete()` and `scale_y_discrete()`. Internally, ggplot2 handles discrete scales by mapping each category to an integer value and then drawing the geom at the corresponding coordinate location. Labels of the discrete variable can be set by `labels` and `guide_axis()` controls their layout.

```{r fig.height=2, fig.width=6}
base <- ggplot(mpg, aes(manufacturer, hwy)) + geom_boxplot()

base + guides(x = guide_axis(n.dodge = 3)) |
# as.character() is not needed for numbers
base + scale_x_discrete(labels = 1:length(unique(mpg$manufacturer)))+
       guides(x = guide_axis(angle = 90))
```

## Binned
Continuous variables can be sliced into multiple bins (e.g. `geom_count`).

```{r fig.height=2, fig.width=6}
base <- ggplot(mpg, aes(hwy, class)) + 
  geom_count()

base |
base + scale_x_binned(n.breaks = 10)
```

## Limits
To make sure different plots share the same axes and legend, just like faceted plots, multiple limits (x, y, color, etc.) should be set.

```{r fig.height=2, fig.width=6}
mpg_99 <- mpg %>% filter(year == 1999)
mpg_08 <- mpg %>% filter(year == 2008)

base_99 <- ggplot(mpg_99, aes(displ, hwy, color = fl)) + 
  geom_point() 
base_08 <- ggplot(mpg_08, aes(displ, hwy, color = fl)) + 
  geom_point()

base_99 +
  scale_x_continuous(limits = c(1, 7)) +
  scale_y_continuous(limits = c(10, 45)) +
  scale_color_discrete(limits = c("c", "d", "e", "p", "r"))|
base_08 + 
  # lims() is another option
  lims(x = c(1, 7), y = c(10, 45), 
       color = c("c", "d", "e", "p", "r"))
```

# Color scales and legends
## Continuous color scales
There are multiple ways to specify continuous color scales. There are many palettes available, but the common interface is **paletteer** package.

```{r fig.height=3, fig.width=4.6}
erupt <- ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_raster() +
  scale_x_continuous(NULL, expand = c(0, 0)) + 
  scale_y_continuous(NULL, expand = c(0, 0)) + 
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line=element_blank(),)

plots <- list(erupt,
              erupt + scale_fill_viridis_c(option = "magma"),
              erupt + scale_fill_distiller(palette = "RdPu"),
              erupt + scico::scale_fill_scico(palette = "lajolla"),
              erupt + scale_fill_fermenter(),
              erupt + paletteer::scale_fill_paletteer_c("viridis::plasma"))
wrap_plots(plots,ncol = 3)
```

The default scale for continuous fill scales is `scale_fill_continuous()` which in turn defaults to `scale_fill_gradient()`. Gradient scales provide a robust method for creating any color scheme you like:

* `scale_fill_gradient()` produces a two-color gradient
* `scale_fill_gradient2()` produces a three-color gradient with specified midpoint
* `scale_fill_gradientn()` produces an n-color gradient

Other packages, such as [**munsell**](https://github.com/cwickham/munsell/) and [**colorspace**](https://colorspace.r-forge.r-project.org/) also provide color palettes.

```{r fig.height=1.5, fig.width=4.6}
# munsell example
erupt + scale_fill_gradient(
  low = munsell::mnsl("5P 2/12"), 
  high = munsell::mnsl("5P 7/12")
) |
# colorspace examples
erupt + 
  scale_fill_gradientn(colors = colorspace::heat_hcl(7))|
erupt + 
  scale_fill_gradientn(colors = colorspace::diverge_hcl(7))
```

Missing values are set to grey by default, which can be changed by set `na.value` option. Breaks can be entirely suppressed by setting them to `NULL`.
```{r fig.height=1.5, fig.width=4.6}
df <- data.frame(x = 1, y = 1:5, z = c(1, 3, 2, NA, 5))
base <- ggplot(df, aes(x, y)) + 
  geom_tile(aes(fill = z), size = 5) + 
  labs(x = NULL, y = NULL)

base |
base + scale_fill_gradient(na.value = NA)|
base + scale_fill_gradient(na.value = "yellow",
                           breaks = NULL)+
        scale_x_continuous(breaks = NULL)
```

## Discrete color scales
The default scale for discrete colors is `scale_fill_discrete()` which in turn defaults to `scale_fill_hue()`, and `scale_color_brewer()` is a discrete color scale that—along with the continuous analog `scale_color_distiller()` and binned analog `scale_color_fermenter()` using handpicked [“ColorBrewer”](http://colorbrewer2.org/) colors.

The default color scheme picks evenly spaced hues around the HCL color wheel. However, it becomes hard to tell the different colors apart when the number of colors exceed eight. Chroma, luminance, and the range of hues, can be controlled with the h, c and l arguments in `scale_fill_hue`.
```{r fig.height=1.5, fig.width=4.6}
df <- data.frame(x = c("a", "b", "c", "d"), 
                 y = c(3, 4, 1, 2))
bars <- ggplot(df, aes(x, y, fill = x)) + 
  geom_bar(stat = "identity") + 
  labs(x = NULL, y = NULL) +
  theme(legend.position = "none")

# Chroma ranging from 0 (grey) to a maximum
# that varies depending on combination of hue and luminance.
bars + scale_fill_hue(c = 40)|
# Hue ranges from 0 to 360 (an angle)
bars + scale_fill_hue(h = c(180, 300))|
# luminance ranging from 0 (black) to 100 (white)
bars + scale_fill_hue(l = 90)
```

It is better to use `scale_fill_grey()`, which maps discrete data to grays, from light to dark, to print graphics in black and white.

You can also use `scale_fill_manual()` to set the colors manually. 

Color scales also come in binned versions. The default scale is `scale_fill_binned()` which in turn defaults to `scale_fill_steps()`. `scale_fill_steps()` is analogous to `scale_fill_gradient()`, and allows you to construct your own two-color gradients.
```{r fig.height=1.5, fig.width=4.6}
erupt + scale_fill_steps(low = "grey", high = "brown")|
erupt + scale_fill_steps2(low = "grey", mid = "white", 
                          high = "brown", midpoint = .02)|
erupt + scale_fill_stepsn(n.breaks = 12, 
                          colors = terrain.colors(12))
```

A brewer analog for binned scales is called `scale_fill_fermenter()`, which does not interpolate between the brewer colors if you set `n.breaks` larger than the number of colors in the palette.

## Alpha scales
Alpha scales map the transparency of a shade to a value in the data. They are not often useful, but can be a convenient way to visually down-weight less important observations. `scale_alpha()` is an alias for `scale_alpha_continuous()`.

## Legends
A number of settings that affect the overall display of the legends are controlled through the `theme` system. 

* `legend.position`: The position and justification of legends ( “right”, “left”, “top”, “bottom”, “none”, or a numeric vector of length two).
* `legend.direction`: layout of items in legends ("horizontal" or "vertical").
* `legend.box`: arrangement of multiple legends ("horizontal" or "vertical").
* `legend.box.just`: justification of each legend within the overall bounding box, when there are multiple legends ("top", "bottom", "left", or "right").
* `legend.margin = unit(0, "mm")`: suppressing margin around the legends.

```{r fig.height=2, fig.width=6}
base <- ggplot(toy, aes(up, up)) + 
  geom_point(aes(color = txt), size = 3)

base + theme(legend.position = c(0, 1), 
             legend.justification = c(0, 1))|
base + theme(legend.position = c(0.5, 0.2), 
             legend.justification = c(1, 0))|
base + theme(legend.position = c(1, 0), 
             legend.justification = c(1, 0))
```

The `key_glyph` argument can be used to associate a particular layer with a different kind of glyph. More precisely, each geom is associated with a function such as `draw_key_path()`, `draw_key_boxplot()`, which can be used to pass desired keys to `key_glyph`.

```{r fig.height=2, fig.width=6}
base <- ggplot(economics, aes(date, psavert, 
                              color = "savings"))+
  theme(legend.position = "bottom")

base + geom_line()|
base + geom_line(key_glyph = "timeseries")|
base + geom_line(key_glyph = draw_key_boxplot) 
```

The legend guide displays individual keys in a table. The most useful options are:

*   `nrow` or `ncol` which specify the dimensions of the table. `byrow` controls how the table is filled: `FALSE` fills it by column (the default), `TRUE` fills it by row.
* `reverse` reverses the order of the keys.
* `override.aes` is useful when you want the elements in the legend display differently to the geoms in the plot.
* `keywidth` and `keyheight` (along with `default.unit`) allow you to specify the size of the keys. These are grid units, e.g. `unit(1, "cm")`.

```{r fig.height=2, fig.width=6}
base <- ggplot(mpg, aes(displ, hwy, color = drv)) +
  geom_point(size = 4, alpha = .2, stroke = 0)

base + guides(color = guide_legend()) |
base + guides(color = guide_legend(override.aes = list(alpha = 1)))
```

`guide_bins()` is suited to the situation when a continuous variable is binned and then mapped to an aesthetic that produces a legend, such as size, color and fill. 

*   `axis` indicates whether the axis should be drawn (default is `TRUE`).
*   `direction` is a character string specifying the direction of the guide.
*   `show.limits` specifies whether tick marks are shown at the ends of the 
    guide axis.
*   `axis.color`, `axis.linewidth` and `axis.arrow` are used to
    control the guide axis that is displayed alongside the legend keys.
*   `keywidth`, `keyheight`, `reverse` and `override.aes` have the same
  behavior as `guide_legend()`.

```{r fig.height=3, fig.width=8}
base <- ggplot(mpg, aes(displ, manufacturer, size = hwy)) +
  geom_point(alpha = .2) +
  scale_size_binned()

base + guides(size = guide_bins(show.limits = TRUE,
                                axis = FALSE))|
base + guides(size = guide_bins(axis.colour = "red",# must be colour
              axis.arrow = arrow(length = unit(.1, "inches"), 
                                 ends = "first", 
                                 type = "closed")))
```

The color bar guide `guide_colorbar()` is designed for **continuous ranges of colors**, which outputs a rectangle where the color gradient varies.  The most important arguments are:

* `barwidth` and `barheight` control the size of the bar, e.g. `unit(1, "cm")`.
* `nbin` controls the number of slices (default value is 20).
* `reverse` flips the color bar.

The `guide_colorsteps()` guide is a version of `guide_colorbar()` appropriate for **binned color and fill scales**. Arguments mostly mirror those for `guide_colorbar()`. The additional arguments are as follows:

* `show.limits` indicates whether values should be shown at the ends of the stepped color bar.
* `ticks` is a logical variable indicating whether tick marks should be displayed (default is `NULL`: inherit from the scale).
* `even.steps` is a logical variable indicating whether bins should be evenly spaced (default is `TRUE`) or proportional in size to their frequency in the data.

```{r fig.height=2, fig.width=6}
base <- ggplot(mpg, aes(displ, hwy, color = cyl)) + 
      geom_point() + 
      scale_color_binned()
base2 <- ggplot(mpg, aes(cyl, displ, color = hwy)) +
  geom_point(size = 2)

base + 
  guides(color = guide_colorsteps(show.limits = TRUE,
                                  ticks = TRUE)) |
base2 + 
  guides(color = guide_colorbar(barheight = unit(2, "cm"),
                                ticks = FALSE))|
base2 + 
  guides(color = guide_colorbar(barwidth = unit(0.5, "cm"),
                                nbin = 5))
```

# Other aesthetics
## Size
The default scale for size aesthetics is `scale_size()` in which a linear increase in the variable is mapped onto a linear increase in the **area**. By default the values in the data (more precisely in the scale limits) is mapped to sizes of 1 to 6, which can be changed by `range` option. To map values to **radius**, `scale_radius()` should be used.

```{r fig.height=3, fig.width=5}
planets <- data.frame(
  name = c("Mercury", "Venus", "Earth", "Mars", 
           "Jupiter", "Saturn", "Uranus", "Neptune"),
  type = c(rep("Inner", 4), rep("Outer", 4)),
  position = 1:8,
  radius = c(2440, 6052, 6378, 3390, 71400, 60330, 25559, 24764),
  orbit = c(57900000, 108200000, 149600000, 227900000, 
            778300000, 1427000000, 2871000000, 4497100000))
planets$name <- with(planets, factor(name, name))
# plot
base <- ggplot(planets, aes(1, name, size = radius)) + 
  geom_point() + 
  scale_x_continuous(breaks = NULL) + 
  labs(x = NULL, y = NULL, size = NULL)
  
base + 
  # default
  scale_size(range = c(1, 6))+
  ggtitle("not to scale")|
base + 
  scale_radius(limits = c(0, NA), range = c(0, 10)) + 
  ggtitle("to scale")
```

## Shape
To merge different aesthetics to one legend, just set the same name.
```{r fig.height=1.5, fig.width=4}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = drv, shape = drv)) + 
  scale_color_discrete("Drive train")+
  scale_shape_discrete("Drive train")
```

## Line type

```{r fig.height=2, fig.width=9}
# define line types
df_lines <- data.frame(
  linetype = factor(
    1:4,
    labels = c("solid", "longdash", "dashed", "dotted")))
# map variable to linetype
base <- ggplot(economics_long, aes(date, value01))
base + geom_line(aes(group = variable)) |
base + geom_line(aes(linetype = variable))|
# See scale_manual for more flexibility
ggplot(df_lines) +
  geom_hline(aes(linetype = linetype, yintercept = 0), 
             size = 2) +
  scale_linetype_identity() +
  facet_grid(linetype ~ .) +
  theme_void(20)
```

## Manual scales
You can create new scale with the “manual” version of each: `scale_linetype_manual()`, `scale_shape_manual()`, `scale_color_manual()`, etc. Valid aesthetic values are described in `vignette("ggplot2-specs")`.

## Identity scales
Identity scales, such as `scale_color_identity()` and `scale_shape_identity()`, are used when your data is already scaled such that the data and aesthetic spaces are the same. 

```{r fig.height=3, fig.width=5}
ggplot(luv_colours, aes(u, v)) + 
geom_point(aes(color = col), size = 3) + 
scale_color_identity() + 
coord_equal()
```
