---
title: 'ggplot2: Layers'
author: Fei
date: '2021-03-29'
slug: ggplot2-layers
categories:
  - R
tags:
  - plot
  - R
weight: 1
draft: no
output:
  blogdown::html_page:
    toc: yes
    toc_depth: 3
    number_sections: yes
---

```{r include=FALSE}
knitr::opts_chunk$set(warning = F, message = F, cache = TRUE,tidy=F,fig.align='center',fig.showtext=TRUE,results="hold",fig.show = "hold")
library(tidyverse)
library(patchwork)
```

# Introduction

In general, there are three purposes for a layer:

* To display the data. 
* To display a statistical summary of the data. 
* To add additional metadata: context, annotations, and references.

# Geoms
Geoms can be roughly divided into individual and collective geoms. 

* __Individual geoms__ draw a distinct graphical object for each observation (row). For example, the point geom draws one point per row.
  * `geom_bar(stat = "identity")` makes a bar chart. We need stat = "identity" because the default stat automatically counts values
  * `geom_area()` draws an area plot, which is a line plot filled to the y-axis (filled lines).
  * `geom_rect()` draw rectangles (parameterised by the four corners of the rectangle).
  * `geom_tile()` also draw rectangles (parameterised by the center of the rect and its size).
  * `geom_raster()` is a fast special case of `geom_tile()` used when all the tiles are the same size.
* __Collective geoms__ display multiple observations with one geometric object.
  * The assignment of observations to graphical elements can be controlled by `group` aesthetic (shown below).
  * If a group is defined by a **combination of multiple variables**, use `interaction()` to combine them, e.g. `aes(group = interaction(school_id, student_id))`.
  * Setting the grouping aesthetic in `ggplot()` will affect all geoms below.
  * Lines and paths operate on a **“first value” principle**: each segment is defined by two observations, and ggplot2 applies the aesthetic value (e.g., color) associated with the first observation when drawing the segment.
  * Mapping aesthetics to discrete variables works well for bar and area plots, resulting in **stacking** individual pieces.

```{r fig.height=2, fig.width=4}
library(babynames)
hadley <- dplyr::filter(babynames, name == "Hadley")
ggplot(hadley, aes(year, n, group = sex, color = sex)) + 
  geom_line()
```

# Statistical summaries

## Revealing uncertainty
There are four basic families of geoms using the aesthetics `ymin` and `ymax` to show uncertainty:

* Discrete x, range: `geom_errorbar()`, `geom_linerange()`
* Discrete x, range & center: `geom_crossbar()`, `geom_pointrange()`
* Continuous x, range:` geom_ribbon()`
* Continuous x, range & center: `geom_smooth(stat = "identity")`

```{r fig.height=2, fig.width=6}
y <- c(18, 11, 16)
df <- data.frame(x = 1:3, y = y, se = c(1.2, 0.5, 1.0))
base <- ggplot(df, aes(x, y, 
                       ymin = y - se, ymax = y + se))
base + geom_crossbar()|
base + geom_pointrange()|
base + geom_smooth(stat = "identity")
```

```{r fig.height=2, fig.width=8}
base + geom_errorbar()|
# if set width of errorbar to 0, it will be same as linerange
base + geom_errorbar(width = 0)|
base + geom_linerange()|
base + geom_ribbon()
```

## Weighted data
Data can be weighted by the `weight` aesthetic.
```{r fig.height=2, fig.width=6}
# Unweighted
ggplot(midwest, aes(percwhite, percbelowpoverty)) + 
  geom_point() + 
  geom_smooth(method = lm, size = 1) |

# Weighted by population
ggplot(midwest, aes(percwhite, percbelowpoverty)) + 
  geom_point(aes(size = poptotal / 1e6)) + 
  geom_smooth(aes(weight = poptotal), 
              method = lm, size = 1) +
  scale_size_area(guide = "none")
```

## Displaying distributions
There are a number of geoms that can be used to display distributions. To compare the distribution between groups, you have a few options: 

* Show small multiples of the histogram, `facet_wrap(~ var)`.
* Use color and a frequency polygon, `geom_freqpoly()`.
* Use a "conditional density plot", `geom_histogram(position = "fill")`.

An alternative to a bin-based visualisation is a density estimate. `geom_density()` places a little normal distribution at each data point and sums up all the curves.
```{r fig.height=6, fig.width=6}
# define layout
layout <- "
AB
CD
EF
"
# Use color and a frequency polygon
ggplot(diamonds, aes(depth)) + 
  geom_freqpoly(aes(color = cut), 
                binwidth = 0.1, na.rm = TRUE) +
  theme(legend.position = "none") +
# Use a "conditional density plot"
ggplot(diamonds, aes(depth)) + 
  geom_histogram(aes(fill = cut), 
                 binwidth = 0.1, position = "fill",
                 na.rm = TRUE) +
  theme(legend.position = "none") +
# density plot
ggplot(diamonds, aes(depth)) +
  geom_density(na.rm = TRUE) + 
  theme(legend.position = "none") +
# adjust binwidth
ggplot(diamonds, aes(depth)) + 
  geom_histogram(aes(fill = cut), 
                 binwidth = 0.2, position = "fill",
                 na.rm = TRUE) +
  theme(legend.position = "none") +
# adjust = 1/2 means use half of the default bandwidth (more rough)
ggplot(diamonds, aes(depth)) +
  geom_density(na.rm = TRUE, adjust = 0.5) + 
  theme(legend.position = "none") +
# both fill and color can separate groups
ggplot(diamonds, aes(depth, fill = cut)) +
  geom_density(alpha = 0.2, na.rm = TRUE) + 
  theme(legend.position = "none") +
# apply layout and set common x axis
plot_layout(design = layout) & 
  scale_x_continuous(limits = c(58, 68))
```

## Dealing with overplotting
When the data is large, points on a scatterplot will overlap. This problem is called overplotting, which can be solved by a number of ways:

* Making points smaller (`geom_point(shape = ".")`) or using hollow glyphs (`geom_point(shape = 1)`).
* Making points transparent (`geom_point(alpha = 1 / 10)`).
* Alleviate some overlaps with `geom_jitter()`, and adjust points by `width` and `height` arguments.
* Breaking the plot into many small squares or hexagons.

```{r fig.height=2, fig.width=6}
df <- data.frame(x = rnorm(2000), 
                 y = rnorm(2000))
norm <- ggplot(df, aes(x, y)) + 
  xlab(NULL) + ylab(NULL)
norm + geom_bin2d() |
# adjust bins
norm + geom_hex(bins = 10)
```

* Adding data summaries to help guide the eye to the true shape of the pattern within the data

## Statistical summaries
Some summary functions like [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html?q=stat%20_%20summary) can produce y, ymin and ymax aesthetics.
```{r fig.height=2, fig.width=8}
# show number of observations in each bin
ggplot(diamonds, aes(color)) + 
  geom_bar() |
# show average price
ggplot(diamonds, aes(color, price)) + 
  geom_bar(stat = "summary_bin", fun = mean) +
  coord_cartesian(ylim=c(2000,6000)) |
# show average price with se
ggplot(diamonds, aes(color, price)) + 
  stat_summary(fun = mean, geom = "bar") +
  stat_summary(fun.data = mean_se, 
               geom = "errorbar",width = 0.1) +
  coord_cartesian(ylim=c(2000,6000)) 
```

## Surfaces
The ggplot2 package does not support true 3d surfaces, but it does support many common tools for summarizing 3d surfaces in 2d. 
```{r fig.height=2, fig.width=8}
small <- faithfuld[seq(1, nrow(faithfuld), by = 10), ]
# contour
# the .. notation (..level..)
# refers to a variable computed internally 
ggplot(faithfuld, aes(eruptions, waiting)) + 
  geom_contour(aes(z = density, color = ..level..)) |
# raster
ggplot(faithfuld, aes(eruptions, waiting)) + 
  geom_raster(aes(fill = density)) |
# Bubble plots work better with fewer observations
ggplot(small, aes(eruptions, waiting)) + 
  geom_point(aes(size = density), alpha = 1/3) + 
  scale_size_area()
```

There is another good example for `geom_tile()` in this website: [Top 50 ggplot2 Visualizations](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html).
```{r fig.height=3, fig.width=4}
var <- mpg$class  # the categorical data 
## Prep data (nothing to change here)
nrows <- 10
df <- expand.grid(y = 1:nrows, x = 1:nrows)
categ_table <- round(table(var) * ((nrows*nrows)/(length(var))))

df$category <- factor(rep(names(categ_table), categ_table))  
# NOTE: if sum(categ_table) is not 100 (i.e. nrows^2), 
# it will need adjustment to make the sum to 100.

## Plot
ggplot(df, aes(x = x, y = y, fill = category)) + 
  # set border color and size
  geom_tile(color = "black", size = 0.5) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), trans = 'reverse') +
  scale_fill_brewer(palette = "Set3") +
  labs(title="Waffle Chart", subtitle="'Class' of vehicles",
       caption="Source: mpg") + 
  theme(panel.border = element_rect(size = 1, fill = NA),
        plot.title = element_text(size = rel(1.2)),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line=element_blank(),
        legend.title = element_blank(),
        legend.position = "right")
```

# Annotations
Conceptually, an annotation supplies metadata for the plot: that is, it provides additional information about the data being displayed.

## Plot and axis titles
The `labs()` helper function lets you to set the various titles using name-value pairs like `title = "My plot title"`. Mathematical expressions wrapped in `quote()`. The rules by which these expressions are interpreted can be found by typing `?plotmath`. To enable markdown you need to set the relevant theme element to `ggtext::element_markdown()`.

```{r fig.height=2, fig.width=4}
base <- ggplot() + xlim(-2*pi, 2*pi)
base + 
  geom_function(fun = rlang::as_function(~ sin(.x^3)))+
  labs(x = "Axis title with *italics* and **boldface**",
       y = quote( f(x) == sin(x^3) ))+
  theme(axis.title.x = ggtext::element_markdown())
```

There are two ways to remove the axis label. Setting `labs(x = "")` omits the label but still allocates space; setting `labs(x = NULL)` removes the label and its space.

## Text labels
The function `geom_text()` adds label text at the specified x and y positions. It has the most aesthetics of any geom, because there are so many ways to control the appearance of a text:

* `family` aesthetic provides the name of a font.
* `fontface` aesthetic specifies the face, which can be “plain” (the default), “bold” or “italic”.
* `hjust` (“left”, “center”, “right”, “inward”, “outward”) and `vjust` (“bottom”, “middle”, “top”, “inward”, “outward”) aesthetics controls alignment of the text.
* `size` aesthetic specifies the font size in millimeters (There are 72.27 pts in a inch, so to convert from points to mm, just multiply by 72.27 / 25.4).
* `angle` controls the rotation of the text in degrees.

In addition to the various aesthetics, `geom_text()` has three parameters:

* `nudge_x` and `nudge_y` are useful to offset the text label.
* If `check_overlap = TRUE`, overlapping labels will be automatically removed from the plot.

Another function `geom_label()` draws a rounded rectangle behind the text.

## Building custom annotations
One useful way to annotate this plot is to use shading to indicate which president was in power at the time. The `annotate()` helper function which creates the data frame.
```{r fig.height=3, fig.width=5}
presidential <- subset(presidential, 
                       start > economics$date[1])
yrng <- range(economics$unemploy)
xrng <- range(economics$date)
caption <- paste(
  strwrap("Unemployment rates in the US have 
  varied a lot over the years", 40), 
  collapse = "\n")

ggplot(economics) + 
  # use geom_rect() to introduce shading
  geom_rect(
    aes(xmin = start, xmax = end, fill = party), 
    ymin = -Inf, ymax = Inf, alpha = 0.2, 
    data = presidential
  ) + 
  # use geom_vline() to introduce separators
  geom_vline(
    aes(xintercept = as.numeric(start)), 
    data = presidential,
    color = "grey50", alpha = 0.5
  ) + 
  # use geom_text to add labels
  geom_text(
    aes(x = start, y = 2500, label = name), 
    data = presidential, 
    size = 3, vjust = 0, hjust = 0, nudge_x = 50
  ) + 
  # use geom_line() to overlay the data on top of 
  # these background elements
  geom_line(aes(date, unemploy)) + 
  scale_fill_manual(values = c("blue", "red")) +
  annotate(geom = "text", 
           x = xrng[1], y = yrng[2], 
           label = caption,
           hjust = 0, vjust = 1, size = 4)+
  # change legend position
  theme(legend.position = "bottom")+
  xlab("date") + 
  ylab("unemployment")
```

Highlight a subset of points can be achieved by drawing larger points in a different color underneath the main data set. `geom_curve()` and `geom_segment()` can be used to draw curves and lines connecting points with labels.

```{r fig.height=1.8, fig.width=3}
p <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(
    data = filter(mpg, manufacturer == "subaru"), 
    color = "orange",
    size = 3) +
  geom_point() 

p + annotate(
  geom = "curve", x = 4, y = 35, xend = 2.65, yend = 27, 
  curvature = .3, arrow = arrow(length = unit(2, "mm"))) +
  annotate(geom = "text", x = 4.1, y = 35, label = "subaru", 
           hjust = "left")
```

# Arranging plots
As shown before, the [patchwork](https://patchwork.data-imaginist.com/) package can arrange plots flexibly.

* `+` does not specify any specific layout, only that the plots should be displayed together.
* layout can be controlled by `plot_layout()`, like `plot_layout(ncol = 2)`.
* operators `/` and `|` can facilitate generating columns and rows.
* operators can be grouped by parentheses (e.g. `p3 | (p2 / (p1 | p4))`).
* complex layouts can be specified in the `design` argument in `plot_layout()`.
* guides can be combined and shown in a particular region (e.g. `p1 + p2 + p3 + guide_area() + plot_layout(ncol = 2, guides = "collect")`).
* `&` can change theme and axis for all the graphics.
* tags can be nested by setting `tag_level` in `plot_layout()`.

```{r eval=FALSE, include=TRUE}
p123[[2]] <- p123[[2]] + plot_layout(tag_level = "new")
p123 + plot_annotation(tag_levels = c("I", "a"))
```

* plots can be shown on top of each other using `inset_element()`.
```{r eval=FALSE, include=TRUE}
p24 <- p2 / p4 + plot_layout(guides = "collect")
p1 + 
  inset_element(
    p24, 
    left = 0.4, 
    bottom = 0.4, 
    right = unit(1, "npc") - unit(15, "mm"), 
    top = unit(1, "npc") - unit(15, "mm"),
    align_to = "full"
  )
```
