---
title: 'ggplot2: Getting started'
author: Fei
date: '2021-03-29'
slug: ggplot2-getting-started
categories:
  - R
tags:
  - plot
  - R
weight: 1
draft: no
output:
  blogdown::html_page:
    toc: yes
    toc_depth: 3
    number_sections: yes
---

```{r include=FALSE}
knitr::opts_chunk$set(warning = F, message = F, cache = TRUE,tidy=F,fig.align='center',fig.showtext=TRUE,results="hold",fig.show = "hold")
library(tidyverse)
library(patchwork)
```

# Introduction
The book *ggplot2: Elegant Graphics for Data Analysis* is a good starting point for learning [ggplot2](https://ggplot2.tidyverse.org/index.html), a useful R package for producing graphics. I summarized some main points and useful tips here. For more details, please refer to [the online version of the book](https://ggplot2-book.org/introduction.html). Packages required for this book are listed below.

```{r eval=FALSE, include=TRUE}
install.packages(c(
  "directlabels", "dplyr", "gameofthrones", "ggforce", "gghighlight", 
  "ggnewscale", "ggplot2", "ggraph", "ggrepel", "ggtext", "ggthemes", 
  "hexbin", "mapproj", "maps", "munsell", "ozmaps", "paletteer", 
  "patchwork", "rmapshaper", "scico", "seriation", "sf", "stars", 
  "tidygraph", "tidyr", "wesanderson" 
))
```

Wilkinson created the grammar of graphics, which is the underlying grammar of ggplot2 (and for which "gg" stands). All plots are composed of the **data**, the information to be visualized, and a **mapping**, the description of relationships between variables and aesthetic attributes, which includes five components:

* **Layer**: collection of geometric elements (**geoms**: points, lines, polygons, etc.) and statistical transformations (**stats**).
* **Scale**: mapping values in the data space to values in the aesthetic space including the use of color, shape or size, as well as drawing the legend and axes(an inverse mapping).
* **Coord**: coordinate system (Cartesian coordinate, polar coordinates), providing axes and gridlines system.
* **Facet**: specifying how to break up and display subsets of data (a.k.a. conditioning or latticing/trellising).
* **Theme**: controling the finer points of display

# Getting started
## Aesthetic attributes
There is one scale for each aesthetic mapping in a plot. In the first plot, the value “blue” is **scaled to a pinkish color**, and a legend is added, because aesthetic mapping of color is set **inside the aes()**. If you want to set an aesthetic to a fixed value, without scaling it, do so in the individual layer **outside of aes()**, which is shown in the second plot.

```{r fig.height=3, fig.width=7}
f1 <- ggplot(mpg, aes(displ, hwy)) + geom_point(aes(color = "blue"))
f2 <- ggplot(mpg, aes(displ, hwy)) + geom_point(color = "blue")
f1 | f2
```

## Faceting
To facet a plot you simply add a faceting specification with `facet_wrap()`. For example, `facet_wrap(~group,ncol = 1,scales="free")` will show different groups in tables of graphics using different scales, arranged in one column.

## Plot geoms
There are many geom functions other than `geom_point()`, here are some examples:

* `geom_smooth()` fits a smoother to the data and displays the smooth and its 
  standard error (can be turned off with `geom_smooth(se = FALSE)`).
  * `method = "loess"`, the default for small n, uses a smooth local 
      regression.
  * `method = "gam"` fits a generalized additive model.
  * `method = "lm"` fits a linear model, giving the line of best fit.
  * `method = "rlm"` works like `lm()`, but uses a robust fitting algorithm so 
    that outliers don't affect the fit as much.
* `geom_boxplot()`, `geom_jitter()`, and `geom_violin()` produce plot to summarize the distribution of a set of points.
* `geom_histogram()` and `geom_freqpoly()` show the distribution of 
  continuous variables.
* `geom_bar()` shows the distribution of categorical variables.
* `geom_path()` and `geom_line()` draw lines between the data points. However, `geom_path` connect points in the order of presentation. 

More functions can be found in the [cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf)

## Modifying the axes
The x- and y-axis labels can be modified by `xlab()` and `ylab()`. The limits of axes are controled by `xlim()` and `ylim()`.

## Output
There are a few things you can do with a plot object:

* Render it on screen with `print()`.
* Save it to disk with `ggsave()`. For example, use `ggsave("plot.png", p, width = 5, height = 5)` to save plot as png.
* Briefly describe its structure with `summary()`.
* Save a cached copy of it to disk, with `saveRDS()`. This saves a complete
    copy of the plot object, so you can easily re-create it with `readRDS()`.
